Bayesian Sparsification Methods for Deep Complex-valued Networks

### 2020-06-23 paper/paper.tex
cite han-learning for mebedded motivation and shoretn certain paragraphs

% general intro text with motivation
Deep neural networks are an integral part of machine learning and data science toolset
for practical data-driven problem solving. With continual miniaturization ever more
applications can be found in embedded systems. Common embedded applications include
on-device image recognition and signal processing. Despite recent advances in generalization
and optimization theory specific to deep networks, deploying in actual embedded hardware
remains a challenge due to storage, real-time throughput, and arithmetic complexity
restrictions \citep{he_amc:_2018}. Therefore, compression methods for achieving high
model sparsity and numerical efficiency without losing much in performance are especially
relevant.
% Solving the storage and artihmetic constraints encourages development of model compression
% and sparsification methods, that focus on favourable trade-off between performance and size.

@@

% overview
In broad terms Bayesian Inference is a principled framework of reasoning about uncertainty
and a tool to update prior beliefs about model's parameters in accordance with evidence
or empirical data into a posterior distribution. The posterior is useful for inference
regarding unobserved data, predictive statistics, parameter confidence regions, and
model's uncertainty.
% intuitive definition of  of model given data is the same as probability of data under hypothesis (model)
% KL-div min & VI -- \citep{jordan_introduction_1999} approximate Bayesian inference
% SVI -- \citep{hoffman_stochastic_2013} scalable stochastic variant
% SGVB -- \citep{kingma_auto-encoding_2014} differentiable MCMC estimator of the evidence lower bound for SVI

@@

% speed, scale, variance, family of suitable Var approximations
In subsequent years several improvements to Variational Inference approach were introduced.
To make VI able to handle large-scale datasets \citet{hoffman_stochastic_2013} proposed
\emph{Stochastic Variational Inference}, which uses stochastic gradient optimization of
\eqref{eq:elbo_general} based on noisy unbiased gradient estimates of ELBO
computed on random mini-batches from the dataset. \citet{titsias_doubly_2014} translated
dependence on $\theta$ from $q_{\theta}$ to the functions inside the expectations of ELBO
and propose \emph{Doubly Stochastic Variational Inference}, which computes unbiased gradient
estimator based on both dataset subsampling and random draws from variational distribution
$q_{\theta}$. Note, that the KL divergence term can replaced by its unbiased differentiable
finite-sample estimator without forfeiting convergence of (doubly) SVI.

@@

% VD is a Bayesian perspective on dropout regularization \citep{hinton_improving_2012} offered
% by \citet{kingma_variational_2015} and repurposed for sparsification by \citet{molchanov_variational_2017}.
%
\citet{kingma_variational_2015} provided a unifying perspective on Dropout, DropConnect
\citep{wan_regularization_2013}, and Gaussian Dropout \citep{wang_fast_2013} through
the lens of Variational Inference and propose \emph{Variational Dropout}. They argued
that the multiplicative noise introduced by Dropout methods induces a distribution
equivalent to a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \eta_j$ with $
  \eta_j \sim p_\theta(\eta_j)
$ iid from some $p_\theta(\eta)$.
%
% Binary dropout with rate $p \in (0, 1)$ uses $
%   p_\eta
%     = \mathcal{B}\bigl(
%       \{0, \tfrac1{1-p}\}, 1-p
%     \bigr)
% $, whereas in Gaussian dropout $
%   \eta \sim \mathcal{N}(1, \alpha)
% $ for $\alpha = \tfrac{p}{1-p}$.
% This suggests making $\alpha$ a variational parameter
% and optimizing it in \eqref{eq:elbo_general} with an appropriate penalty term. Thus,

@@

Due factorization assumption, the term $KL(q_\theta \| \pi)$ in \eqref{eq:sgvb_estimator}
for Sparse VD unravels into $
  \sum_j K(\tfrac{\sigma^2_{j}}{\mu_{j}^2})
$ with
\begin{equation}  \label{eq:improper-kl-div-real}
  K(\alpha)
    \propto \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
\citet{kingma_variational_2015} approximated $K(\alpha)$ over $\alpha \in (0, 1)$ by a
polynomial with a logarithmic term, and later \citet{molchanov_variational_2017} refined
the approximation of \eqref{eq:improper-kl-div-real} by weighted sum of a sigmoid and a
soft-plus term.
% see p.6 sec 3.4 of kingma_variational_2015
In appendix~\ref{sec:real-chisq-grad} we verify the value and the derivative of their
approximation against a Monte-Carlo estimate for $\alpha$ varying over a fine logarithmic
grid and the exact expression for gradient of \eqref{eq:improper-kl-div-real}.
% (!) good for SGVB: gradient-based methods require unbiased gradient estimates
% (in) it is in poor taste to brag about this beign the first time anyone has derived a
% derivative of an intractable function.

@@

For a fully factorized approximation $q(\omega)$ and factorized prior $
  \pi(\omega) = \prod_{ij} \pi(\omega_{ij})
$, the divergence term in ELBO \eqref{eq:sgvb_estimator} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q \| \pi)
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = - \sum_{ij}
        \mathbb{H}(q(\omega_{ij}))
        + \mathbb{E}_{q(\omega_{ij})} \log{\pi(\omega_{ij})}
    \,.
\end{equation}
We consider two fully factorized priors: an improper prior, resembling VD, and $\cplx$-Gaussian
ARD prior. We omit subscripts ${ij}$ for brevity in this section.

@@

% The training consists of three successive stages: ``pre-train'' $\to$ ``sparsify'' $\to$ ``fine-tune''.
At the \emph{``pre-train''} stage every network is fit ``as-is'', retaining its original
deterministic architecture and using only the likelihood term from \eqref{eq:sgvb_estimator}.

During the \emph{``sparsify''} stage we make every layer Bayesian and apply $\cplx$-valued
variational sparsification (sec.~\ref{ssub:vd_prior}, ~\ref{ssub:ard_prior}) or its $\real$
counterpart. We inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:sgvb_estimator}:
\begin{equation}  \label{eq:elbo_with_coef}
  - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
  + \frac1{M} \sum_{k=1}^M
      \log p_{\phi}(x_{i_k} \mid g(\varepsilon_k; \theta))
  \,.
  \tag{\ref{eq:sgvb_estimator}'}
\end{equation}
In contrast to \cite{molchanov_variational_2017}, who anneal $C$ from zero to one during
training, we use constant $C$ and vary it between runs. This allows us to explore the
compression-performance profile by balancing model's likelihood and posterior's penalty
for diverging form the sparsifying prior in \eqref{eq:elbo_with_coef}. In particular,
higher $C$ implies higher sparsity.
% RFC: citation needed fot this "trick". Cite $\beta$-VAE maybe?

@@

The \emph{``fine-tune''} stage reverts the network back to deterministic layers,
initializes the parameters to the mode of $q_\theta(\omega)$ and proceeds as the
``pre-train'' stage, except for training only those parameters, which are specified
by sparsity masks.

@@

Having verified the $\cplx$ variational sparsification method on MNIST-like datasets
and simple models, we turn to the CIFAR10 dataset comprising $32\times 32$ colour
images of $10$ classes \citep{krizhevsky_learning_2009} and focus on the VGG16 network
\citep{simonyan_very_2015}.
%
% same experimental setup as in MNIST, describe the VGG16 model and its cplx-valued modification
We train the VGG16 network and its $\cplx$ variant, in which we have replaced $\real$-valued
layers with their $\cplx$-valued counterparts. We do not halve or double the features
in any network, since the goal of this experiment is to assess the trade-off for a deep
convolutional network.
%
% differences with mnist
Unlike experiment in sec.~\ref{sub:mnist_like_datasets}, we consider the raw features
only, use full training split and allocate $20$, $40$, and $20$ epochs to each stage.
During training every mini-batch of $128$ samples is augmented by random horizontal flipping
and random cropping, which is done by zero-padding the image with four pixels and extracting
a $32\times 32$ patch from the $40\times 40$ intermediate image.
%
We vary $C$ in \eqref{eq:elbo_with_coef} over $
  \{\tfrac32 2^{-\tfrac{k}2} \colon k=7, \cdots, 15\}
$, compare $\real / \cplx$ VD and ARD methods and measure accuracy on CIFAR10 test split.

@@

% effects of extra training epochs and the impact of tau
We take the models compressed with $C \in \{\tfrac1{20}, \frac1{200}\}$ and re-run only
the fine-tuning stage for various pruning thresholds $
  \tau \in \{\tfrac{k}2\colon k=-8\cdots+8\}
$. The performance-compression curves depicted in figure~\ref{fig:hist__and__threshold__tradeoff}
are parameterized by decreasing $\tau$ from left to right, since models are not re-compressed
which makes $\tau$ monotonically affect the compression rate. From \eqref{eq:elbo_with_coef}
and the relative positions of the curves it can be concluded that $C$ has a much more
substantial impact on the compression profile of each method, than the choice of the
pruning threshold.
% coarser grid than in top plot in order not to wait for 20 compute days, but 9.5 on
% 4 x Geforce 1080Ti, 256 Gb, core-i7 machine, running 4*2 experiments in parallel

@@

% why is there such a dramamtic difference in impact of ``fine-tune''?
We provide the following interpretation of the apparent contrast in performance impact
borne by fine-tuning between less than $\times50$ and higher than $\times100$ compression
regimes in figure~\ref{fig:musicnet__trade-off}, also observed in sec.~\ref{sub:mnist_like_datasets}.
%
% * early stopping kick in at around 10-20 epoch
% * performance peaks at 10-15 epoch and then gradually declines
The value of $C$ in \eqref{eq:elbo_with_coef} is a good proxy for the ranking of the
final compression rate since it directly affects the feedback from sparsifying prior.
So, during the $50$ epoch allotted for ``sparsify'' stage, low $C$ prevents the sparsity
inducing prior from pulling the posterior sufficiently away from the likelihood-maximizing
parameters inherited from the ``pre-train'' stage. It is reasonable, therefore, to expect
that for undercompressed models the fine-tuning stage acts essentially as a continuation
of pre-training. And, since we have observed that longer training invariably deteriorates
the validation performance, the ``fine-tune'' stage should lead to overfitting for small
$C$.
%
Figure~\ref{fig:musicnet__early_stopping} shows that the models, which have been
sparsified with $C$ less than $\tfrac1{400}$ and have $50\times$ compression, need
considerably less training epochs before early stopping terminates the process.

\begin{figure}[!t]
  \centering
  \begin{subfigure}[b]{1.\columnwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=\columnwidth]{figure__fine-tune_fx__early__compression.pdf}
  \end{subfigure}
  \caption{%
    Early stopping epoch at fine-tuning stage.
  }
  \label{fig:musicnet__early_stopping}
\end{figure}

@@

The results of this study have direct implications for real-time signal processing with
embedded deep learning applications both in terms of lower storage requirements and higher
throughput stemming from fewer floating-point multiplications, despite higher arithmetic
complexity of $\cplx$-valued networks.

>>>>

% general intro text with motivation
Deep neural networks are an integral part of machine learning and data science toolset
for practical data-driven problem solving. With continual miniaturization ever more
applications can be found in embedded systems. Common embedded applications include
on-device image recognition and signal processing. Despite recent advances in generalization
and optimization theory specific to deep networks, deploying in actual embedded hardware
remains a challenge due to storage, real-time throughput, and arithmetic complexity
restrictions \citep{han_learning_2015}. Therefore, compression methods for achieving high
model sparsity and numerical efficiency without losing much in performance are especially
relevant.
% Solving the storage and artihmetic constraints encourages development of model compression
% and sparsification methods, that focus on favourable trade-off between performance and size.

@@

% overview
In broad terms Bayesian Inference is a principled framework for reasoning about uncertainty
and updating prior beliefs about model's parameters in accordance with evidence or
empirical data into a posterior distribution. The posterior is useful for inference
regarding unobserved data, predictive statistics, parameter confidence regions, and
model's uncertainty.
% intuitive definition of  of model given data is the same as probability of data under hypothesis (model)
% KL-div min & VI -- \citep{jordan_introduction_1999} approximate Bayesian inference
% SVI -- \citep{hoffman_stochastic_2013} scalable stochastic variant
% SGVB -- \citep{kingma_auto-encoding_2014} differentiable MCMC estimator of the evidence lower bound for SVI

@@

% speed, scale, variance, family of suitable Var approximations
In subsequent years several improvements to Variational Inference approach were introduced.
To make VI able to handle large-scale datasets \citet{hoffman_stochastic_2013} proposed
\emph{Stochastic Variational Inference}, which uses stochastic gradient optimization of
\eqref{eq:elbo_general} based on noisy unbiased gradient estimates of ELBO
computed on random mini-batches from the dataset. \citet{titsias_doubly_2014} translated
the dependence on location–scale parameters of $q_{\theta}$ to the function inside its
expectation and proposed \emph{Doubly Stochastic Variational Inference}. DSVI constructs
an unbiased finite-sample estimator of the gradient of \eqref{eq:elbo_general} by both
subsampling the dataset and sampling from $q_{\theta}$, without forfeiting convergence
of SVI.

@@

% VD is a Bayesian perspective on dropout regularization \citep{hinton_improving_2012} offered
% by \citet{kingma_variational_2015} and repurposed for sparsification by \citet{molchanov_variational_2017}.
%
\citet{kingma_variational_2015} provided a unifying perspective on Dropout, DropConnect
\citep{wan_regularization_2013}, and Gaussian Dropout \citep{wang_fast_2013} through
the lens of Variational Inference and propose \emph{Variational Dropout}. They argued
that the multiplicative noise introduced by Dropout methods induces a distribution
equivalent to a fully factorized variational posterior $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \eta_j$ with $
  \eta_j \sim p_\theta(\eta_j)
$ iid from some $p_\theta(\eta)$.
%
% Binary dropout with rate $p \in (0, 1)$ uses $
%   p_\eta
%     = \mathcal{B}\bigl(
%       \{0, \tfrac1{1-p}\}, 1-p
%     \bigr)
% $, whereas in Gaussian dropout $
%   \eta \sim \mathcal{N}(1, \alpha)
% $ for $\alpha = \tfrac{p}{1-p}$.
% This suggests making $\alpha$ a variational parameter
% and optimizing it in \eqref{eq:elbo_general} with an appropriate penalty term. Thus,

@@

Due factorization assumption, the term $KL(q_\theta \| \pi)$ in \eqref{eq:sgvb_estimator}
for Sparse VD unravels into $
  \sum_j K(\tfrac{\sigma^2_{j}}{\mu_{j}^2})
$ with
\begin{equation}  \label{eq:improper-kl-div-real}
  K(\alpha)
    \propto \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
\citet{kingma_variational_2015} approximated $K(\alpha)$ over $\alpha \in (0, 1)$ by a
polynomial with a logarithmic term, and later \citet{molchanov_variational_2017} refined
the approximation of \eqref{eq:improper-kl-div-real} by weighted sum of a sigmoid and a
soft-plus term.
% see p.6 sec 3.4 of kingma_variational_2015
In appendix~\ref{sec:real-chisq-grad} we verify the derivative of their approximation
against a Monte-Carlo estimate for $\alpha$ varying over a fine $\log$-scale grid and
the exact expression for gradient of \eqref{eq:improper-kl-div-real}.
% (!) good for SGVB: gradient-based methods require unbiased gradient estimates
% (in) it is in poor taste to brag about this beign the first time anyone has derived a
% derivative of an intractable function.

@@

For a fully factorized approximation $q(\omega)$ and factorized prior $
  \pi(\omega) = \prod_{ij} \pi(\omega_{ij})
$, the divergence term \eqref{eq:sgvb_estimator} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q \| \pi)
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = - \sum_{ij}
        \mathbb{H}(q(\omega_{ij}))
        + \mathbb{E}_{q(\omega_{ij})} \log{\pi(\omega_{ij})}
    \,.
\end{equation}
We consider two fully factorized priors: an improper prior, resembling VD, and $\cplx$-Gaussian
ARD prior. We omit subscripts ${ij}$ for brevity in this section.

@@

% The training consists of three successive stages: ``pre-train'' $\to$ ``sparsify'' $\to$ ``fine-tune''.
At the \emph{``pre-train''} stage every network is fit ``as-is'' using deterministic layers
and only the likelihood term from \eqref{eq:sgvb_estimator}.

During the \emph{``sparsify''} stage we make every layer stochastic and apply $\cplx$-valued
variational sparsification (sec.~\ref{ssub:vd_prior}, ~\ref{ssub:ard_prior}) or its $\real$
counterpart. We inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:sgvb_estimator}:
\begin{equation}  \label{eq:elbo_with_coef}
  - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
  + \frac1{M} \sum_{k=1}^M
      \log p_{\phi}(x_{i_k} \mid g(\varepsilon_k; \theta))
  \,.
  \tag{\ref{eq:sgvb_estimator}'}
\end{equation}
In contrast to \cite{molchanov_variational_2017}, who anneal $C$ from zero to one during
training, we use constant $C$ and vary it between runs. This allows us to explore the
compression-performance profile by balancing model's likelihood and posterior's penalty
for diverging form the sparsifying prior in \eqref{eq:elbo_with_coef}. In particular,
higher $C$ implies higher sparsity.
% RFC: citation needed fot this "trick". Cite $\beta$-VAE maybe?

@@

At the \emph{``fine-tune''} stage the network reverts back to deterministic architecture
and proceeds the same way as the ``pre-train'' stage, except for training only those
parameters, which are specified by sparsity masks.

@@

Having verified the $\cplx$ variational sparsification method on MNIST-like datasets
and simple models, we turn to the CIFAR10 dataset comprising $32\times 32$ colour
images of $10$ classes \citep{krizhevsky_learning_2009} and focus on the VGG16 network
\citep{simonyan_very_2015}.
%
% same experimental setup as in MNIST, describe the VGG16 model and its cplx-valued modification
We train the VGG16 network and its $\cplx$ variant, in which we have replaced $\real$-valued
layers with their $\cplx$-valued counterparts. We do not halve or double the features
in any network, since the goal of this experiment is to assess the trade-off for a deep
convolutional network.
%
% differences with mnist
Unlike experiment in sec.~\ref{sub:mnist_like_datasets}, we consider the raw features
only, use full training split and allocate $20$, $40$, and $20$ epochs to each stage.
During training every mini-batch of $128$ samples is augmented by random horizontal flipping
and random cropping, which is done by zero-padding the image with four pixels and extracting
a $32\times 32$ patch from the $40\times 40$ intermediate image.
%
We vary $C$ in \eqref{eq:elbo_with_coef} over $
  \{\tfrac32 2^{-\tfrac{k}2} \colon k=7, \cdots, 15\}
$, compare $\real / \cplx$ VD and ARD methods and measure accuracy on the usual test split.

@@

% effects of extra training epochs and the impact of tau
We take the full models compressed with $C \in \{\tfrac1{20}, \frac1{200}\}$ and re-run
only the fine-tuning stage for various pruning thresholds $
  \tau \in \{\tfrac{k}2\colon k=-8\cdots+8\}
$. The performance-compression curves depicted in figure~\ref{fig:hist__and__threshold__tradeoff}
are parameterized by decreasing $\tau$ from left to right, since models are not re-compressed
which makes $\tau$ monotonically affect the compression rate. From \eqref{eq:elbo_with_coef}
and the relative positions of the curves it can be concluded that $C$ has a much more
substantial impact on the compression profile of each method, than the choice of the
pruning threshold.
% coarser grid than in top plot in order not to wait for 20 compute days, but 9.5 on
% 4 x Geforce 1080Ti, 256 Gb, core-i7 machine, running 4*2 experiments in parallel

@@

% why is there such a dramamtic difference in impact of ``fine-tune''?
We provide the following interpretation of the apparent contrast in performance impact
borne by fine-tuning between less than $\times50$ and higher than $\times100$ compression
regimes in figure~\ref{fig:musicnet__trade-off}, also observed in sec.~\ref{sub:mnist_like_datasets}.
%
% * early stopping kick in at around 10-20 epoch
% * performance peaks at 10-15 epoch and then gradually declines
The value of $C$ in \eqref{eq:elbo_with_coef} is a good proxy for the ranking of the
final compression rate since it directly affects the feedback from sparsifying prior.
So, during the $50$ epoch allotted for ``sparsify'' stage, low $C$ prevents the sparsity
inducing prior from pulling the posterior sufficiently away from the likelihood-maximizing
parameters inherited from the ``pre-train'' stage. It is reasonable, therefore, to expect
that for undercompressed models the fine-tuning stage acts essentially as a continuation
of pre-training. And, since we have observed that longer training invariably deteriorates
the validation performance, the ``fine-tune'' stage should lead to overfitting for small
$C$.
%
Figure~\ref{fig:musicnet__early_stopping} shows that the models, which have been sparsified
with $C$ less than $\tfrac1{400}$, have less than $50\times$ compression and need considerably
less training epochs before early stopping terminates the process.

\begin{figure}[!t]
  \centering
  \begin{subfigure}[b]{1.\columnwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=\columnwidth]{figure__fine-tune_fx__early.pdf}
  \end{subfigure}
  \caption{%
    Early stopping epoch at fine-tuning stage.
  }
  \label{fig:musicnet__early_stopping}
\end{figure}

@@

The results of this study have direct implications for real-time signal processing with
embedded deep learning applications both in terms of lower storage requirements and higher
throughput stemming from fewer floating-point multiplications, despite higher arithmetic
complexity of $\cplx$-valued networks.
% CVNN capacity grows by \sqrt{2}, compute grows by 2 % RVNN caption by 2 compute by 2.
